---
output:
  word_document: default
  pdf_document: default
  html_document: default
---
# Final Project

```{r}
library(imputeTS)
library(forecast)
library(zoo)
library(ggplot2)
```


### Read in CO2 data

```{r}
co2 <- read.csv("monthly co2.csv")
```

### Explore data

```{r}
str(co2)
```

```{r}
head(co2)
```

### Trim to CO2 column and data since 1966

```{r}
co2 <- co2[,c("Year", "Month", "average")]
colnames(co2) <- c("Year", "Month", "CO2")
co2 <- co2[co2$Year >= 1966, ]
co2.ts <- ts(co2$CO2, start = 1966, frequency = 12)
head(co2.ts)
```

### Plot time series
Observe any missing values or outliers

```{r}
#plotNA.distribution(co2.ts)
plot.ts(co2.ts, main = "Atmostpheric CO2 Over Time", ylab = "Parts Per Million by Volume")
ggplot(co2, aes(x=Year, y=CO2)) + 
  geom_line() + 
  labs(y = "CO2 Parts Per Million by Volume", title = "Atmospheric CO2 Over Time")
```

Data < 0 for atmospheric CO2 doesn't make sense so we will remove these and impute data.

```{r}
co2.ts <- ifelse(co2.ts < 0, NA, co2.ts)
```

```{r}
plotNA.distribution(co2.ts)
```

# Impute missing values

```{r}
co2.ts.imp <- na.interpolation(co2.ts)
```

```{r}
plotNA.distribution(co2.ts.imp)
```

_Conclusions:_ Plot of CO2 time series indicates positive, additive trend and annual seasonality.

### ACF

```{r}
#acf(co2.ts.imp, lag = 100)
# figure
# subplot(2,1,1)
# autocorr(co2.ts.imp)
# subplot(2,1,2)
# parcorr(co2.ts.imp)

par(mfrow=c(2,1))
plot(acf(co2.ts.imp, plot=FALSE), ylab = "CO2 Autocorrelation", main = "")
pacf(co2.ts.imp)
```
Long memory auto correlation. The persistence of high values in acf
plot indicate a long term positive trend. These results are consistent with our plot of the time series.

### PACF

```{r}
pacf(co2.ts.imp,  main = "", ylab = "CO2 Partial Autocorrelation")
```
Long memory partial autocorrelation drops off around lag 20.

_why doesn't x adjust with lag?_

### Training and test
```{r}
co2.ts.train <- ts(co2.ts.imp[1:588], start = 1966, frequency = 12) # through 2014
co2.ts.test <- ts(co2.ts.imp[589:600], start = 2015, frequency = 12) # all of 2015
```

### Seasonal ARIMA
Seasonal time series

```{r}
sarima.mod <- auto.arima(co2.ts.train) # model through 2014
summary(sarima.mod)
```

### Seasonal ARIMA Forecast

```{r}
(sarima.2015 <- forecast(sarima.mod, h=12))
```

```{r}
plot(co2.ts.imp[1:600])
lines(sarima.2015$mean, col = "green")
```

_Calculate error..._

### Holt-Winters
Time series with positive trend and seasonality

```{r}
hw.co2.mod <- HoltWinters(co2.ts.train)
summary(hw.co2.mod)
```

```{r}
hw.co2.mod
```

```{r}
plot(hw.co2.mod)
```

```{r}
plot(fitted(hw.co2.mod))
```

```{r}
# forecast
```


### Read in annual temps

```{r}
temps <- read.csv("GlobalTemperatures.csv")
#temps.ho <- read.csv("16-17_temps.csv")
```

```{r}
names(temps)
```

```{r}
temps <- temps[,c("dt", "LandAndOceanAverageTemperature")]
temps$dt <- as.character(temps$dt)
```


Grab only data since 1966 

```{r}
temps <- temps[temps$dt>= "1966-01-01",]
colnames(temps) <- c("dt", "temp")
```

Streamline holdout data - 2015

```{r}
# temps.ho$dt <- as.Date(as.yearmon(paste(temps.ho$year, temps.ho$month, "01", sep = "-"))) # get date format
# temps.ho$dt <- as.character(temps.ho$dt) # match type in temps df
# temps.ho <-temps.ho[,c("dt", "raw.temp")]
# colnames(temps.ho) <- c("dt", "temp")
# temps.full <- rbind(temps, temps.ho) # make a full list
# temps.full.ts <- ts(temps.full$temp, start = 1966, frequency = 12)

temps.ts <- ts(temps$temp, start = 1966, frequency = 12)
temps.train <- temps[temps$dt<= "2014-12-01",]
temps.test <- temps[temps$dt >= "2015-01-01",]
```


### Make Training Time Series

```{r}
temps.train.ts <- ts(temps.train$temp, start = 1966, frequency = 12)
```


### Plot missing values

```{r}
plotNA.distribution(temps.ts, main = "Global Temperature Over Time", ylab = "Temperature (°C)")
```




```{r}
acf(temps$temp, main = "Temperature Autocorrelation")
pacf(temps$temp, main = "Temperature Partial Autocorrelation", lag = 50) # these are better
```

### Plot both TS together

```{r}
# library(grid)
# library(dplyr)
# 
# co2$dt <- as.Date(as.yearmon(paste(co2$Year, co2$Month, "01", sep = "-"))) # get dt column in co2
# plot1 <- co2 %>%
#   select(dt, CO2) %>%
#   na.omit() %>%
#   ggplot() + geom_line(aes(x = dt, y = CO2)) + ylab("Atmospheric CO2 ppm") + 
#   theme_minimal() +
#   theme(axis.title.x = element_blank())
# 
# plot2 <- temps %>%
#   select(dt, temp) %>%
#   ggplot() + geom_line(aes(x = dt, y = temp)) + ylab("Temperature (°C)") +
#   theme_minimal() +
#   theme(axis.title.x = element_blank())
# 
# grid.newpage()
# grid.draw(rbind(ggplotGrob(plot1), ggplotGrob(plot2), size = "last"))

##qplot(temps.ts) + geom_line(y=temps.ts)

#ggplot(temps, aes(x=dt, y=temp)) + geom_line()
```


### Seasonal ARIMA
Seasonal time series

```{r}
sarima.temps.mod <- auto.arima(temps.train.ts) # model through 2015
summary(sarima.temps.mod)
```

### Seasonal ARIMA Forecast

```{r}
(sarima.temps.2015 <- forecast(sarima.temps.mod, h=12))
```

```{r}
plot(sarima.temps.2015, xlab = "Time", ylab = "Temperature (°C)", main = "Global Temperature ARIMA Forecast")
lines(temps.ts)
```


```{r}
plot(temps.ts)
lines(sarima.temps.2015$mean, col = "green")
```

### Analyze Residuals

```{r}
plot(sarima.temps.mod$residuals)
```

```{r}
acf(sarima.temps.mod$residuals)
```


### Create function to calculate symmetric mean absolute percentage error (sMAPE) for forecast evaluation

```{r}
sMAPE <- function(actual, estimate) {
    absDev <- abs(estimate - actual)
    return(sum(absDev/(estimate + actual))/length(actual))
}
```

```{r}
(sMAPE.sarima.temps <- sMAPE(temps.test$temp, sarima.temps.2015$mean))
```


### Holt-Winters
Time series with positive trend and seasonality

```{r}
hw.temps.mod <- HoltWinters(temps.train.ts)
summary(hw.temps.mod)
```

```{r}
hw.temps.mod
```

```{r}
plot(hw.temps.mod)
```

```{r}
plot(hw.temps.mod$fitted)
```

## Analyzing Residuals

```{r}
plot(hw.temps.mod$fitted[,1]-temps.train.ts)
```
Residuals seem like white noise

```{r}
acf(hw.temps.mod$fitted[,1]-temps.train.ts)
```
No residual autocorrelation

```{r}
(hw.temps.2015 <- forecast(hw.temps.mod, h=12))
```

```{r}
plot(hw.temps.2015, main = "Global Temperature Holt-Winters Forecast", xlab = "Time", ylab = "Temperature (°C)")
lines(ts(temps$temp))
```


### sMAPE Calculation

```{r}
(sMAPE.hw.2015 <- sMAPE(temps.test$temp, hw.temps.2015$mean))
```








